Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-10 manuscript/main.tex
maximizer now has hat

	The goal is to find a unit-norm direction $\xi \in \mathcal{H}$, whose functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, that maximizes the empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \xi \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \xi \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$.
	We denote the solution as
	\begin{equation}
		\hat{\xi} 
		= 
		(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}),
	\end{equation}
	where $\hat{\boldsymbol{\zeta}} \in \mathbb{R}^p$ is the scalar part, and each functional part $\hat{\xi}_j \in L^2([0,1])$ lies in $\operatorname{span}(b_1(t), \ldots, b_M(t))$ and is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) := 
		\sum_{m=1}^M d_{jm} \, b_m(t), \quad j = 1, \ldots, K.
	\end{equation*}

	Let $\mathbf{d}_j := (d_{j1}, \ldots, d_{jM})^\top \in \mathbb{R}^M$ and $\mathbf{y} := (y_1, \ldots, y_n)^\top$. 
	Let $B \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions, with entries
	\[
	B_{m, m'} := \langle b_m, b_{m'} \rangle, \quad m, m' = 1, \ldots, M.
	\]
	We compute $\mathbf{d}_1, \ldots, \mathbf{d}_K$ and $\hat{\boldsymbol{\zeta}}$ as follows.
	First, we obtain the unnormalized coefficient vectors. For each $j$, the unnormalized vector $\bar{\mathbf{d}}_j$ is computed by solving
	\begin{equation*}
		B \bar{\mathbf{d}}_j = (\Theta_j B)^\top \mathbf{y},
	\end{equation*}
	The unnormalized scalar component is given by
	\begin{equation}
		\bar{\boldsymbol{\zeta}} := Z^\top \mathbf{y}.
	\end{equation}
	Next, we normalize all components with respect to the hybrid Hilbert norm (see Definition~\ref{def:hilbert_space}) to ensure $\hat{\xi}$ has unit length. Specifically, we define
	\begin{equation}\label{def:computed_PLS_direction}
		\mathbf{d}_j := 
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\mathbf{d}}_j, \quad j = 1, \ldots, K,
	\end{equation}
	and
	\begin{equation}
		\hat{\boldsymbol{\zeta}} :=
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\boldsymbol{\zeta}}.
	\end{equation}
	Details of the hybrid norm computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.

@@

	\subsection{Proof of new Proposition}
	For any $\xi = (\xi_1, \ldots, \xi_K, \boldsymbol{\zeta})\in \mathcal{H}$, we have:
		\widehat{\textnormal{Cov}} (
		\langle \widetilde{W}, \xi  \rangle_{\mathcal{H}}, Y 
		) 
		&=  %1
		\frac{1}{n} 
		\sum 
		\limits_{i=1}^n 
		y_i
		\langle \widetilde{W}_i, \xi \rangle_\mathcal{H} \\
		&= %2
		\frac{1}{n} 
		\sum \limits_{i=1}^n
		y_i
		\biggl(
		\sum_{j=1}^K
		\langle \widetilde{X}_{ij}, \xi_j \rangle  
		+
	 \mathbf{Z}_i^\top 
		\boldsymbol{\zeta} 
		\biggr)
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\\&= %3
		\frac{1}{n}
		\sum
		\limits_{i=1}^n 
		y_i
		\biggl( \sum \limits_{k=1}^K \int_0^1 \widetilde{X}_{ij}(t) \, \xi_j(t) \, dt \biggr)
		+
		 \frac{1}{n}
		 \sum
		 \limits_{i=1}^n
		 y_i
		 ( \mathbf{Z}_i^\top \boldsymbol{\zeta} )
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		  \\&=  
		  \frac{1}{n} \sum \limits_{i=1}^n 
		  y_i
		\biggl\{ 
		  \sum \limits_{k=1}^K \int_0^1
		   \biggl(
		   \sum_{m=1}^M \theta_{ijm} b_{m}(t) 
		   \biggr) 
		   \biggl(
		   \sum_{m'=1}^M d_{jm'} b_{m'}(t) 
		   \biggr) 
		   dt
		\biggr\}  
		  +
	\frac{1}{n}
	\mathbf{y}^\top
	Z \boldsymbol{\zeta} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\frac{1}{n} \sum \limits_{i=1}^n 
y_i
\biggl\{ 
\sum \limits_{k=1}^K 
\sum_{m=1}^M
\sum_{m'=1}^M  
\theta_{ijm}
d_{jm'}
\int_0^1
 b_{m}(t) 
 b_{m'}(t) 
dt
\biggr\}  
+
\frac{1}{n}
\mathbf{y}^\top
Z \boldsymbol{\zeta} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\sum \limits_{k=1}^K 
\biggl\{
\frac{1}{n} \sum \limits_{i=1}^n 
y_i
(
\boldsymbol{\theta}_{ij}^\top
B
\mathbf{d}_j
)  
\biggr\}
+
\frac{1}{n}
\mathbf{y}^\top
Z \boldsymbol{\zeta}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\frac{1}{n}
\biggl(
\sum \limits_{k=1}^K 
\mathbf{y}^\top
\Theta_{j}
B
\mathbf{d}_j
+
\mathbf{y}^\top
Z \boldsymbol{\zeta} 
\biggr).
		\jongmin{
		Another way of computing \eqref{def:eigenproblem_smoothed}, rather than solving an eigenproblem, is as follows. As shown in the proof in Appendix \ref{section:proof:eigenproblem_derivation}, the empirical covariance is simplified as:
			\\&=
			\frac{1}{n}\tilde{\mathbf{y}}^\top
			\begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix} \;      \underbrace{
				\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
			}_{:= \boldsymbol{\beta}}.
		Let us 
		denote $\mathbf{b} =  \widetilde{\mathbf{y}}^\top \begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix}$. This is the weighted row sum of $\begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix}$, weighted by $\widetilde{\mathbf{y}}$.
		Then using the symmetricity of $J^\ast$, we solve a linear system $J^\ast \mathbf{d} =
		\mathbf{b}^\top$ to obtain $\mathbf{d}$. Then we have
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\widehat{\textnormal{Cov}} (\langle \mathbf{W}, \xi \rangle_{\mathcal{H}}, Y ) 
			&= 
			%%%%%%%%%%%%%%%%%%%%%%%%
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\frac{1}{n} \mathbf{d}^\top J^\ast  \boldsymbol{\beta}
			\\& \overset{(i)}{\leq} 
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\frac{1}{n} \sqrt{\mathbf{d}^\top J^\ast  \mathbf{d}}
			\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}}
			\\& \overset{(ii)}{=}
			\frac{1}{n} \sqrt{\mathbf{d}^\top J^\ast  \mathbf{d}}    ,
		where step $(i)$ uses the Cauchy-Schwarz inequality with the inner product $\langle x, y \rangle_{J^\ast} = x^\top J^\ast y$,
		and step $(ii)$ uses the norm constraint.
		By the Cauchy-Schwarz inequality, the maximum displayed in the last term is obtained by a unit vector (with respect to $\langle \cdot, \cdot \rangle_{J^\ast}$) parallel to $\mathbf{d}$. Therefore we have
			\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
			= \frac{1}{\sqrt{\mathbf{d}^\top  J^\ast \mathbf{d} }}\mathbf{d},
		which only involves solving one linear equation.
	}

>>>>

	The goal is to find a unit-norm direction $\xi \in \mathcal{H}$, whose functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, that maximizes the empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \xi \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \xi \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$.
	We denote the solution as
	\begin{equation}
		\hat{\xi} 
		= 
		(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}),
	\end{equation}
	where $\hat{\boldsymbol{\zeta}} \in \mathbb{R}^p$ is the scalar part, and each functional part $\hat{\xi}_j \in L^2([0,1])$ lies in $\operatorname{span}(b_1(t), \ldots, b_M(t))$ and is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) := 
		\sum_{m=1}^M \hat{d}_{jm} \, b_m(t), \quad j = 1, \ldots, K.
	\end{equation*}

	Let $\hat{\mathbf{d}}_j := (\hat{d}_{j1}, \ldots, \hat{d}_{jM})^\top \in \mathbb{R}^M$ and $\mathbf{y} := (y_1, \ldots, y_n)^\top$. 
	Let $B \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions, with entries
	\[
	B_{m, m'} := \langle b_m, b_{m'} \rangle, \quad m, m' = 1, \ldots, M.
	\]
	We compute $\hat{\mathbf{d}}_1, \ldots, \hat{\mathbf{d}}_K$ and $\hat{\boldsymbol{\zeta}}$ as follows.
	First, we obtain the unnormalized coefficient vectors. For each $j$, the unnormalized vector $\hat{\mathbf{d}}_j'$ is computed by solving a linear equation:
	\begin{equation*}
		B \hat{\mathbf{d}}_j' = (\Theta_j B)^\top \mathbf{y},
	\end{equation*}
	The unnormalized scalar component is given by
	\begin{equation}
		\hat{ \boldsymbol{\zeta} }' := Z^\top \mathbf{y}.
	\end{equation}
	Next, we normalize all components with respect to the hybrid Hilbert norm (see Definition~\ref{def:hilbert_space}) to ensure $\hat{\xi}$ has unit length. Specifically, we define
	\begin{equation}\label{def:computed_PLS_direction}
		\hat{ \mathbf{d}}_j := 
		\frac{1}{
			\left(
			\sum_{j=1}^K \hat{\mathbf{d}}_j'^\top B \hat{\mathbf{d}}_j' +
		(\hat{ \boldsymbol{\zeta} }')^\top \hat{ \boldsymbol{\zeta} }'
			\right)^{1/2}
		} \hat{\mathbf{d}}_j', \quad j = 1, \ldots, K,
	\end{equation}
	and
	\begin{equation}
		\hat{\boldsymbol{\zeta}} :=
		\frac{1}{
			\left(
			\sum_{j=1}^K \hat{\mathbf{d}}_j'^\top B \hat{\mathbf{d}}_j' +
		(\hat{ \boldsymbol{\zeta} }')^\top \hat{ \boldsymbol{\zeta} }'
			\right)^{1/2}
		} 	\hat{ \boldsymbol{\zeta} }'.
	\end{equation}
	Details of the hybrid norm computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.

@@

	\subsection{Proof of new Proposition}
	For any $\xi = (\xi_1, \ldots, \xi_K, \boldsymbol{\zeta})\in \mathcal{H}$, we have:
	\begin{align*}
		\widehat{\textnormal{Cov}} (
		\langle \widetilde{W}, \xi  \rangle_{\mathcal{H}}, Y 
		) 
		&=  %1
		\frac{1}{n} 
		\sum 
		\limits_{i=1}^n 
		y_i
		\langle \widetilde{W}_i, \xi \rangle_\mathcal{H} \\
		&= %2
		\frac{1}{n} 
		\sum \limits_{i=1}^n
		y_i
		\biggl(
		\sum_{j=1}^K
		\langle \widetilde{X}_{ij}, \xi_j \rangle  
		+
	 \mathbf{Z}_i^\top 
		\boldsymbol{\zeta} 
		\biggr)
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		\\&= %3
		\frac{1}{n}
		\sum
		\limits_{i=1}^n 
		y_i
		\biggl( \sum \limits_{k=1}^K \int_0^1 \widetilde{X}_{ij}(t) \, \xi_j(t) \, dt \biggr)
		+
		 \frac{1}{n}
		 \sum
		 \limits_{i=1}^n
		 y_i
		 ( \mathbf{Z}_i^\top \boldsymbol{\zeta} )
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
		  \\&=  
		  \frac{1}{n} \sum \limits_{i=1}^n 
		  y_i
		\biggl\{ 
		  \sum \limits_{k=1}^K \int_0^1
		   \biggl(
		   \sum_{m=1}^M \theta_{ijm} b_{m}(t) 
		   \biggr) 
		   \biggl(
		   \sum_{m'=1}^M d_{jm'} b_{m'}(t) 
		   \biggr) 
		   dt
		\biggr\}  
		  +
	\frac{1}{n}
	\mathbf{y}^\top
	Z \boldsymbol{\zeta} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\frac{1}{n} \sum \limits_{i=1}^n 
y_i
\biggl\{ 
\sum \limits_{k=1}^K 
\sum_{m=1}^M
\sum_{m'=1}^M  
\theta_{ijm}
d_{jm'}
\int_0^1
 b_{m}(t) 
 b_{m'}(t) 
dt
\biggr\}  
+
\frac{1}{n}
\mathbf{y}^\top
Z \boldsymbol{\zeta} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\sum \limits_{k=1}^K 
\biggl\{
\frac{1}{n} \sum \limits_{i=1}^n 
y_i
(
\boldsymbol{\theta}_{ij}^\top
B
\mathbf{d}_j
)  
\biggr\}
+
\frac{1}{n}
\mathbf{y}^\top
Z \boldsymbol{\zeta}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=  
\frac{1}{n}
\biggl(
\sum \limits_{k=1}^K 
\mathbf{y}^\top
\Theta_{j}
B
\mathbf{d}_j
+
\mathbf{y}^\top
Z \boldsymbol{\zeta} 
\biggr).
	\end{align*}

The squared hybrid Hilbert norm $\langle {\xi},{\xi} \rangle_{\mathcal{H}}$ 
is computed as follows:
\begin{align*}
\langle {\xi},{\xi} \rangle_{\mathcal{H}}
&=
\sum_{j=1}^K
\langle
\xi_j, \xi_j
\rangle
+
\boldsymbol{\zeta}^\top \boldsymbol{\zeta}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=
\sum \limits_{k=1}^K \int_0^1
\biggl(
\sum_{m=1}^M d_{jm'} b_{m}(t) 
\biggr) 
\biggl(
\sum_{m'=1}^M d_{jm'} b_{m'}(t) 
\biggr) 
dt
+
\boldsymbol{\zeta}^\top \boldsymbol{\zeta}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&=
aaaaa
\end{align*}
admits the following matrix representation:
	\[
	\langle \boldsymbol{\xi}, \boldsymbol{\xi} \rangle_{\mathcal{H}} = \sum_{j=1}^K \mathbf{d}_j^\top B \mathbf{d}_j + \boldsymbol{\zeta}^\top \boldsymbol{\zeta}.
	\]

	By the definition of the inner product on $\mathcal{H} = \mathbb{R}^{KM} \times \mathbb{R}^p$, we have
	\[
	\langle \boldsymbol{\xi}, \boldsymbol{\xi} \rangle_{\mathcal{H}} = \sum_{j=1}^K \langle \xi_j, \xi_j \rangle_{L^2([0,1])} + \|\boldsymbol{\zeta}\|^2.
	\]
	Since $\xi_j(t) = \sum_{m=1}^M d_{jm} b_m(t)$, it follows that
	\[
	\langle \xi_j, \xi_j \rangle_{L^2([0,1])} 
	= \int_0^1 \left( \sum_{m=1}^M d_{jm} b_m(t) \right)^2 dt 
	= \mathbf{d}_j^\top B \mathbf{d}_j,
	\]
	where $B$ is the Gram matrix of the basis. Substituting this into the definition yields
	\[
	\langle \boldsymbol{\xi}, \boldsymbol{\xi} \rangle_{\mathcal{H}} = \sum_{j=1}^K \mathbf{d}_j^\top B \mathbf{d}_j + \boldsymbol{\zeta}^\top \boldsymbol{\zeta}.
	\]

		\jongmin{
		Another way of computing \eqref{def:eigenproblem_smoothed}, rather than solving an eigenproblem, is as follows. As shown in the proof in Appendix \ref{section:proof:eigenproblem_derivation}, the empirical covariance is simplified as:
		\begin{align*}
			\widehat{\textnormal{Cov}} (\langle \mathbf{W}, \xi \rangle_{\mathcal{H}}, Y ) &= \frac{1}{n} \sum \limits_{i=1}^n \langle \mathbf{W}_i, \xi \rangle_\mathcal{H} Y_i \\
			&= n^{-1} \widetilde{\mathbf{y}}^\top (\widetilde{C}J\mathbf{a} + \mathbf{Z} \boldsymbol{\theta}).
			\\&=
			\frac{1}{n}\tilde{\mathbf{y}}^\top
			\begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix} \;      \underbrace{
				\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
			}_{:= \boldsymbol{\beta}}.
		\end{align*}
		Let us 
		denote $\mathbf{b} =  \widetilde{\mathbf{y}}^\top \begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix}$. This is the weighted row sum of $\begin{bmatrix} \widetilde{C}J \; \mathbf{Z} \end{bmatrix}$, weighted by $\widetilde{\mathbf{y}}$.
		Then using the symmetricity of $J^\ast$, we solve a linear system $J^\ast \mathbf{d} =
		\mathbf{b}^\top$ to obtain $\mathbf{d}$. Then we have
		\begin{align*}
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\widehat{\textnormal{Cov}} (\langle \mathbf{W}, \xi \rangle_{\mathcal{H}}, Y ) 
			%%%%%%%%%%%%%%%%%%%%%
			&= 
			%%%%%%%%%%%%%%%%%%%%%%%%
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\frac{1}{n} \mathbf{d}^\top J^\ast  \boldsymbol{\beta}
			\\& \overset{(i)}{\leq} 
			\max_{\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}} = 1}
			\frac{1}{n} \sqrt{\mathbf{d}^\top J^\ast  \mathbf{d}}
			\sqrt{\boldsymbol{\beta} J^\ast  \boldsymbol{\beta}}
			\\& \overset{(ii)}{=}
			\frac{1}{n} \sqrt{\mathbf{d}^\top J^\ast  \mathbf{d}}    ,
		\end{align*}
		where step $(i)$ uses the Cauchy-Schwarz inequality with the inner product $\langle x, y \rangle_{J^\ast} = x^\top J^\ast y$,
		and step $(ii)$ uses the norm constraint.
		By the Cauchy-Schwarz inequality, the maximum displayed in the last term is obtained by a unit vector (with respect to $\langle \cdot, \cdot \rangle_{J^\ast}$) parallel to $\mathbf{d}$. Therefore we have
		\begin{equation*}
			\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
			= \frac{1}{\sqrt{\mathbf{d}^\top  J^\ast \mathbf{d} }}\mathbf{d},
		\end{equation*}
		which only involves solving one linear equation.
	}
